---
title: "UN report"
author: "Claire Volk"
date: "2023-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
In the curly brackets, you can name the chunk. For instance, the first code block is named “setup”, and include=FALSE prevents code and results from showing up in the output file.


```{r packages}
library(tidyverse)
```
This report was prepared to the attention of the United Nations. It analyzes the relationship between a country's GDP, its life expectancy and CO2 emissions. Our goal is to determine to what degree a country’s economic strength or weakness may be related to its public health status and impact on climate pollution. We hypothesize that both life expectancy and CO2 emissions increase with a country's GDP.

```{r data}
gapminder_1997 <- read_csv("data/gapminder_1997.csv")
```

```{r gdp_lifeexp_1997}
ggplot(data = gapminder_1997) + 

  aes(x = gdpPercap, y = lifeExp, color=continent, size=pop/1000000) +

  geom_point() +

  labs(x = "GDP Per Capita", y = "Life Expectancy",

       title= "Do people in wealthy countries live longer?", size="Population (in millions)")
```