---
title: "UN report"
author: "Claire Volk"
date: "2023-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#In the curly brackets, you can name the chunk. For instance, the first code block is named “setup”, and include=FALSE prevents code and results from showing up in the output file while echo=FALSE to tell this code chunk that we don’t want to see the code but just the output.
```



```{r packages, include = FALSE}
library(tidyverse)
library(knitr)
```
### Summary of Report
This report was prepared to the attention of the United Nations. It analyzes the relationship between a country's GDP, its life expectancy and CO2 emissions. Our goal is to determine to what degree a country’s economic strength or weakness may be related to its public health status and impact on climate pollution. We hypothesize that both life expectancy and CO2 emissions increase with a country's GDP.

```{r data, include=FALSE}
gapminder_1997 <- read_csv("../data/gapminder_1997.csv")


nCountries <- gapminder_1997 %>%
  select(country) %>%
  n_distinct()

minGDP <- gapminder_1997 %>%
  summarise(round(min(gdpPercap))) %>%
  pull()

maxGDP <- gapminder_1997 %>%
  summarise(round(max(gdpPercap))) %>%
  pull()

#to reference these variables, include backticks:  `r some_R_variable_name `
```
### Plotting the Relationship between GDP per Capita and Life Expectance
```{r gdp_lifeexp_1997, echo = FALSE}
ggplot(data = gapminder_1997) + 

  aes(x = gdpPercap, y = lifeExp, color=continent, size=pop/1000000) +

  geom_point() +

  labs(x = "GDP Per Capita", y = "Life Expectancy",

       title= "Do people in wealthy countries live longer?", size="Population (in millions)")
```

The above plot shows the relationship between GDP per capita and life expectancy
for a total of `r nCountries ` countries. For this set of countries,
economic wealth ranged from a minimum of USD `r minGDP`
to a maximum of USD `r maxGDP` per capita.

### Tables with Useful Information


|Summary of Data|
|------|------|
|Number of Countries|`r nCountries`|
|Minimum GDP per capita|`r minGDP`|
|Maximum GDP per capita|`r maxGDP`|


```{r Kable Table, echo=FALSE}
gapminder_1997 %>% 
  filter(continent == "Oceania") %>% 
  kable()
```
### My Relationship with R
* R is my favorite programming language
* _mutate_ is the most useful function in R although I also like _filter_
* ***I still need to learn how to clean data and do computational microbiology***
